Appendix S2: Forward algorithm description

### 2018-10-21 appendix_s2.Rmd
Formatting for final submission

---
title: Disease and climate effects on individuals drive post-reintroduction
  population dynamics of an endangered amphibian
bibliography: library.bib
output:
    pdf_document:
      keep_tex: true
      includes:
          in_header: header.sty
fontsize: 12pt
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, include = FALSE)
library("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
library(lubridate)
library(tidyverse)
n_digit <- 3
df2num <- function(df, n_digit = 3) {
  stopifnot(nrow(df) == 1)
  stopifnot(ncol(df) == 1)
  df %>%
    unlist %>%
    round(n_digit)
}
```

\newpage

\linenumbers

\begin{abstract}
The emergence of novel pathogens often has dramatic negative effects on previously unexposed host populations. 
Subsequent disease can drive populations and even species to extinction. 
After establishment in populations, pathogens can continue to affect host dynamics, influencing the success or failure of species recovery efforts. 
However, quantifying the effect of pathogens on host populations in the wild is challenging because individual hosts and their pathogens are difficult to observe. 
Here we use long-term mark-recapture data to describe the dynamics of reintroduced populations of an endangered amphibian (\textit{Rana sierrae}) and evaluate the success of these recovery efforts in the presence of a recently-emerged pathogen, the amphibian chytrid fungus \textit{Batrachochytrium dendrobatidis}. 
We find that high \textit{B. dendrobatidis} infection intensities are associated with increases in frog detectability and reductions in survival. 
When average infection intensities are high, adults are more likely to gain infections and less likely to lose infections.
We also find evidence for intensity-dependent survival, with heavily infected individuals suffering higher mortality. 
These results highlight the need in disease ecology for probabilistic approaches that account for uncertainty in infection intensity using imperfect observational data. 
Such approaches can advance the understanding of disease impacts on host population dynamics, and in the current study will improve the effectiveness of species conservation actions.

\bigskip
\noindent \textbf{Key words}: \textit{Batrachochytrium dendrobatidis}; host-pathogen dynamics; mark-recapture; mountain yellow-legged frog; population establishment; recruitment; reintroduction; survival.

\end{abstract}

# Introduction

Amphibians are one of the most threatened groups of vertebrates [@wake2008we]. 
Although the drivers of amphibian decline vary taxonomically and spatially, the amphibian chytrid fungus \textit{Batrachochytrium dendrobatidis} (Bd) is a major cause of population declines and species extinctions in montane habitats worldwide [@fisher2009global; @grant2016quantitative; @skerratt2007spread]. 
In the face of these declines, species recovery will often require introductions to restore extirpated populations, but little is known about the dynamics of population establishment and persistence of threatened amphibians [@armstrong2008directions], especially in the presence of disease. 
Given Bd's role in eliminating populations, we expect introduction outcomes to be shaped by disease impacts on demographic rates, but measuring such impacts in wild populations is difficult [@mccallum1995detecting; @briggs2010enzootic].

The mountain yellow-legged frog is emblematic of global amphibian declines. 
Although formerly abundant in the relatively protected habitats of California's Sierra Nevada mountains (USA), the two species that make up this taxon (\textit{Rana muscosa} and \textit{Rana sierrae} [@vredenburg2007concordant]) have disappeared from over 93\% of historical localities, due primarily to the introduction of nonnative fish into fishless habitats [@knapp2000non; @knapp2005effects] and the emergence of Bd [@vredenburg2010dynamics]. 
In response, both species are listed as ``endangered'' under the U.S. Endangered Species Act and included on the International Union for Conservation of Nature Red List of Threatened Species [@iucn2017]. 
\textit{R. sierrae} has shown recent signs of recovery in the best-protected portion of its range, including Yosemite National Park, but due to dispersal limitations the re-establishment of populations may require introductions in addition to natural recovery [@knapp2016large]. 
The effectiveness of such actions remains unclear, and the drivers of post-introduction population dynamics and introduction success or failure are poorly understood.

Introduction outcomes may be described in terms of survival, recruitment, and abundance trajectories. 
Mark-recapture studies provide a method to estimate these quantities in wild populations while accounting for factors that can complicate inference, including imperfect detection [@jolly1965explicit; @pradel1996utilization], multiple classes within the population that may be imperfectly resolved [@lebreton2002multistate; @conn2009multistate], and incompletely observed individual-level traits [@royle2009analysis]. 
This flexibility is critical for understanding disease impacts in wild populations. 
Multi-state mark-recapture methods have been applied with great success to understand the population-level impacts of Bd, and differences in demographic rates between infected and uninfected classes [@murray2009impact; @pilliod2010effects; @sapsford2015infection; @hudson2016situ].

Assuming that survival depends only on whether an individual is infected, and not considering infection intensity (i.e., load), could be problematic given that Bd-caused frog mortality is often a function of load [@briggs2010enzootic]. 
However, individual-level infection data are often not available. 
For example, Bd load may not be observed if an individual is not encountered or captured on a survey, or if they are captured but no infection data are collected. 
This presents a major challenge: infection intensity is hard to measure, but may be critical for understanding population dynamics. 
Previous efforts to understand disease impacts in wild populations as a function of load at the individual level have randomly imputed missing infection load data with the observed distribution of load values [@spitzen2017fragile]. 
However, random imputation could bias inference if load data are not missing at random, for example, if heavily infected individuals are easier or more difficult to detect than individuals with lower loads.

In this paper, we seek to understand whether introduction outcomes can be explained by post-introduction disease dynamics or climatic conditions. 
We expected that failed introductions might be attributed to some combination of disease dynamics that inhibit population growth and climatic conditions that do not favor adult survival or recruitment.
We use a Bayesian multi-state model with data from decade-long mark-recapture studies in two reintroduced frog populations.
We find that Bd affects introduction success via impacts on adult survival.
Winter severity also appears to regulate adult survival and recruitment.
In addition, we find that infected adults are easier to detect, which raises broader questions about understanding disease impacts in wild populations, and highlights the importance of accounting for hidden individual-level infection dynamics. 
These insights will inform ongoing and future efforts aimed at restoring the  endangered mountain yellow-legged frog, and provide a means to quantitatively assess why some introduction efforts fail and others succeed.

# Methods

## Field sites and methods

The two study lakes are located in eastern Yosemite National Park (California, USA) at elevations of 2880 and 3200 m. \textit{R. sierrae} populations in both lakes were established via translocation from a single nearby donor population (elevation: 3176 m). 
The donor population has been Bd-positive for at least two decades [@fellers2001oral], and contains one of the largest \textit{R. sierrae} populations in Yosemite (minimum population size during this study: 600-1500 adults). 
The introduction lakes are located 5.5-7.0 km from the donor population and both contain high quality \textit{R. sierrae} habitat (i.e., fishless, 6-12 m deep, with adjacent meadow and stream habitats) [@knapp2005effects]. 
Because these lakes harbor a sensitive species, we refer to them by the pseudonyms Alpine and Subalpine [@lindenmayer2017not]. 
At Alpine in the year 2001, five years prior to the translocation of adults into this lake, a single large adult R. sierrae was detected on a visual encounter survey [@knapp2005effects]. 
No \textit{R. sierrae} of any life stage was detected at Alpine on a visual encounter survey conducted three weeks prior to translocation in 2006. 
At Subalpine, a single \textit{R. sierrae} tadpole was observed on a visual encounter survey in 2001, seven years prior to adult translocations in 2008. 
Subsequent surveys in 2005 and 2007 did not detect \textit{R. sierrae} of any life stage at Subalpine.
Based on these observations and the widespread historical occurrence of \textit{R. sierrae} across similar lakes in Yosemite, we assumed that both lakes were historically occupied by \textit{R. sierrae}, and that these populations were extirpated prior to this study's translocations by the introduction of nonnative fish prior to 1977 and the arrival of Bd in the 1980s or 1990s based on historical declines of \textit{R. sierrae} observed at that time [@phillips1990have; @sherman1993population; @drost1996collapse].

Prior to translocations, the presence or absence of Bd at these sites was uncertain. 
It was thought that Bd was likely present at both Alpine and Subalpine, as Bd was ubiquitous or nearly so in Yosemite at that time [@knapp2011nowhere]. 
Both sites and/or adjacent habitats contained other amphibians (Alpine lake: Yosemite toad \textit{Anaxyrus canorus}; Subalpine lake: \textit{A. canorus} and Pacific treefrog \textit{Hyliola regilla}), but the infection status of those species was unknown. 
It was decided in collaboration with Yosemite National Park that reintroduced \textit{R. sierrae} populations would need to persist in the presence of Bd, because preventing individuals from becoming infected was not feasible, and there was no known means to eliminate Bd from the environment. 
The potential risk associated with introducing Bd or a novel strain of Bd to these sites was determined to be low relative to the benefits associated with re-establishing populations of \textit{R. sierrae}.

The initial introduction of frogs to Alpine and Subalpine occurred during summer 2006 and 2008, respectively, and additional introductions to supplement both populations were conducted in subsequent years. 
Prior to introduction, adult frogs ($\geq$ 40 mm snout-vent length (SVL)) were captured at the donor site, and tagged (8mm passive integrated transponder (PIT) tags), measured, and weighed. 
To estimate Bd load, we also collected a skin swab from each frog using standard methods [@Hyatt2007diagnostic; @vredenburg2010dynamics]. 
In the first introduction to Alpine, frogs were transported on foot; in all subsequent introductions to both Alpine and Subalpine, frogs were transported via helicopter. 
The number of individuals introduced to each study site on each introduction date is provided in Table \ref{tab:table1}.

To describe the dynamics of the introduced populations, both populations were assessed for 10-12 years using mark-recapture methods.
Between 2006 and 2012, lakes were visited approximately once per month during the summer active season (June-September) and on a single day (primary period) all habitats were searched repeatedly for frogs which were captured using hand-held nets. Adult frogs were identified via their PIT tag (or tagged if they were untagged), measured, weighed, swabbed, and released at the capture location. 
During 2013-2017, we adopted a robust design in which all habitats were searched during several consecutive days (surveys), and frogs processed as described above. 
Within a primary period, frogs that were captured on more than one survey were measured, weighed, and swabbed only when first captured.

Skin swabs were analysed using standard Bd DNA extraction and qPCR methods [@boyle2004rapid] except that swab extracts were analyzed singly instead of in triplicate [@kriger2006cost, @vredenburg2010dynamics]. 
During 2005-2014, we used standards developed from known concentrations of zoospores [@boyle2004rapid] and after 2014 we used standards based on single ITS1 PCR-amplicons [@longo2013its1]. 
Based on paired comparisons between samples analyzed using both types of standards, Bd in the study area has an average of 60 ITS1 copies per zoospore. 
To express all qPCR results as the number of ITS1 copies, starting quantities obtained using the zoospore standard (measured as “zoospore equivalents”) were multiplied by 60. In addition, all qPCR quantities (regardless of standard) were multiplied by 80 to account for the fact that DNA extracts from swabs were diluted 80-fold during extraction and PCR [@vredenburg2010dynamics].

We acquired hourly air temperature and daily snow depth data from two nearby meteorological stations (ERY and DAN, respectively) and snow water equivalent data from a manually-measured snow course (DAN; California Data Exchange Center - http://cdec.water.ca.gov). 
The stations and snow course are 5-17 km from the study lakes. 
We used snow water equivalent as measured on April 1 of each year as a measure of winter severity, to be used as a covariate for frog recruitment and survival. 
Daily snow depth data were acquired to model survey occurrence as described below.
To better understand the detection process, we averaged hourly air temperature data collected between 0900-1600 on each day a survey took place, to derive a ``survey air temperature'' metric.

## Model development

We developed a hierarchical Bayesian hidden Markov model to understand how environmental factors and partly observed time-varying individual traits (Bd loads) jointly drive population dynamics. 
This model also needed to account for the partly deterministic recruitment process that arises with introductions. 
The result is an extended open population Jolly-Seber mark-recapture model with known additions to the population (introduced adults), continuous uncertain infection states of individuals which may or may not be captured, and sampling that is unevenly distributed in time.

```{r get-pct-missing-swabs}
load('alpine.RData')
alpine_n_missing <- sum(Y == 4)
alpine_n <- sum(Y > 1)

load('subalpine.RData')
subalpine_n_missing <- sum(Y == 4)
subalpine_n <- sum(Y > 1)

total_swabs <- alpine_n + subalpine_n
n_missing <- alpine_n_missing + subalpine_n_missing
prop_miss <- (n_missing) / total_swabs
pct_miss <- 100 * prop_miss
```

In a small number of cases, adults were captured, weighed, and measured, but no swab data were associated with the capture (e.g., if a swab was lost). 
In these cases (`r n_missing` of `r total_swabs` swabs), capture data indicate that the adult was alive during the survey, but its infection status is unknown. 
As such, there are three possible observations for captured frogs: detected/infected, detected/uninfected, and detected/unknown infection status. 
A fourth possible observation class corresponds to non-detection: an individual was not observed during a survey.

## Mark-recapture model structure

We consider four possible observations for each individual $i=1, ..., M$ on each survey $j=1, ..., n_j$: $o_{i, j} = 1$ indicates that the individual was not detected; $o_{i, j}=2$ indicates that the individual was observed and the swab collected for that individual indicated that they were uninfected; $o_{i, j}=3$ indicates that the individual was observed and results from the collected swab showed that the frog had a non-zero Bd load; and $o_{i, j}=4$ indicates that the individual was observed, but no swab data were available from the capture.

We use parameter-expanded data augmentation to account for the fact that the total number of adults in the population is unknown [@royle2012parameter]. 
Across the entire time period of the study, we assume $N_s$ individuals have existed in the population, comprising the superpopulation of individuals that were present at some time. 
While $N_s$ is not directly observable, across all surveys $n \leq N_s$ unique individuals were known to be present in the population (because they were introduced and/or captured). 
An estimate of $N_s$ can be acquired by considering a large number $M > N_s$ of individuals, $M - N_s$ of which never existed [@royle2009analysis]. 
Thus, the model has a state corresponding to individuals that have not recruited yet. 
Here, $M$ was chosen to be 2218 at Alpine (618 observed unique individuals plus 1600 augmented individuals), and 558 at Subalpine (158 observed individuals, and 400 augmented individuals). 
These values were chosen to be considerably greater than our prior guess of $N_s$, and we also verified that posterior estimates of $N_s$ were much less than $M$ to avoid problems on the boundary of this augmented parameter space [@dennis2015computational].

We denote the true state of individual $i$ in primary period $t$ as $u_{i, t}$, for every individual $i = 1, ..., M$ and each primary period $t = 1,..., n_t$. 
The four states that we consider are: $u_{i, t} = 1$ for individuals that have not recruited, $u_{i, t} =2$ for uninfected adults, $u_{i, t} = 3$ for infected adults, and $u_{i, t} = 4$ for dead individuals. 
Each survey $j=1, ..., n_j$ occurs in one of the $n_t$ primary periods, and we denote the primary period in which survey $j$ takes place as $t_j$. 
Each primary period $t$ occurs within one year, but within a year there are multiple primary periods.
We set the year containing the first primary period to $y_{t = 1} = 1$, and generally $y_t$ represents the year containing primary period $t$. 
Years increment by one until the final year of the mark recapture efforts, which we denote $n_y$: $y \in \{1, 2, ..., n_y\}$.
We assume that within a primary period, the state of each individual does not change (i.e., individuals do not recruit into the adult population, gain or lose Bd infection, or die). 
This assumption is justified by the short time intervals between surveys within primary periods.

### Observation model

An emission matrix $\matr{\Omega}_{i, j}$ links observations to hidden states, with the elements of $\matr{\Omega}_{i, j}$ providing the probability of each possible observation of individual $i$ in survey $j$, given the true hidden state. 
The rows in $\matr{\Omega}_{i, j}$ correspond to the state of individual $i$ in primary period $t_j$, and the columns correspond to an observation of individual $i$ in survey $j$:

\[
\matr{\Omega}_{i, j} =
\begin{blockarray}{ccccc}
  \text{Not detected} & \text{Bd negative} & \text{Bd positive} & \text{Bd unknown} & \\
    o_{i, j} = 1 & o_{i, j} = 2 & o_{i, j} = 3 & o_{i, j} = 4 \\
\begin{block}{(cccc)c}
  1 & 0 & 0 & 0 & \begin{tabular}{@{}c@{}}Not recruited \\ $u_{i, t_j} = 1$ \end{tabular} \\
  1 - p_{i, j}^- & p^-_{i, j} \delta & 0 & p^-_{i, j} (1 - \delta) & \begin{tabular}{@{}c@{}}Uninfected \\ $u_{i, t_j} = 2$ \end{tabular} \\
  1 - p_{i, j}^+ & 0 & p^+_{i, j} \delta & p^+_{i, j} (1 - \delta) & \begin{tabular}{@{}c@{}}Infected \\ $u_{i, t_j} = 3$ \end{tabular} \\
  1 & 0 & 0 & 0 & \begin{tabular}{@{}c@{}}Dead \\ $u_{i, t_j} = 4$ \end{tabular} \\
\end{block}
\end{blockarray}
 \]

\noindent The structure of $\matr{\Omega}_{i, j}$ implies that there are no mistaken individual identifications (those that are dead or not recruited are never detected), and that a swab successfully makes it to the lab and provides qPCR data with probability $\delta$, conditional on the animal being detected. 
Detection probabilities are provided by $p^-_{i, j}$ for uninfected (Bd negative) and $p^+_{i, j}$ for infected (Bd positive) individuals, and these detection probabilities can vary by individual and survey.
We also assume that there are no false-positive or false-negative Bd results [@Hyatt2007diagnostic], though relaxing this assumption may be a promising future area, given the potential sensitivity of swab results to infection intensity [@miller2012estimating].

### State model

The hidden states of each individual evolve as a Markov process with transition matrix $\matr{\Psi}_{i, t}$, the entries of which provide the probability of transitioning to state $u_{i, t + 1}$ (the column index) from state $u_{i, t}$ (the row index). 
This matrix is different for individuals that naturally recruit versus those that recruit deterministically due to introduction. 
For naturally recruiting individuals:

\[
\matr{\Psi}_{i, t} =
\begin{blockarray}{ccccc}
  \text{Not recruited} & \text{Uninfected} & \text{Infected} & \text{Dead} & \\
    u_{i, t+1} = 1 & u_{i, t+1} = 2 & u_{i, t+1} = 3 & u_{i, t+1} = 4 \\
\begin{block}{(cccc)c}
  1 - \lambda_t & \lambda_t (1 - \gamma_t) & \lambda_t \gamma_t & 0 & \begin{tabular}{@{}c@{}}Not recruited \\ $u_{i, t} = 1$ \end{tabular} \\
  0 & \phi^-_{i, t} (1 - \eta^+_t) & \phi^-_{i, t} \eta^+_t & 1 - \phi^-_{i, t} & \begin{tabular}{@{}c@{}}Uninfected \\ $u_{i, t} = 2$ \end{tabular} \\
  0 & \phi^+_{i, t} \eta^-_t & \phi^+_{i, t} (1 - \eta^-_t) & 1 - \phi^+_{i, t} & \begin{tabular}{@{}c@{}}Infected \\ $u_{i, t} = 3$ \end{tabular} \\
  0 & 0 & 0 & 1 & \begin{tabular}{@{}c@{}}Dead \\ $u_{i, t} = 4$ \end{tabular} \\
\end{block}
\end{blockarray}
 \]

\noindent where $\lambda_t$ is the probability that an individual enters the adult population between primary periods $t$ and $t + 1$, $\gamma_t$ is the probability that a recruiting individual is infected conditional on entry, $\phi^-_{i, t}$ is a survival probability for an uninfected adult, $\phi^+_{i, t}$ is a survival probability for an infected adult, $\eta^+_t$ is the probability of transitioning from the uninfected to infected class conditional on survival, and $\eta^-_t$ is the probability of transitioning from the infected to uninfected class conditional on survival.

For introduced individuals, the recruitment process is deterministic. 
Specifically, for each introduced adult, there is zero chance that they recruit prior to the primary period in which they are introduced, and if they are introduced at time $t\textsuperscript{intro}_i$, then the probability that they recruit into a particular class (their state upon introduction) must be one. 
For these introductions, all introduced adults were infected, and thus recruited into the infected adult class, leading to the following transition matrix for introduced animals, where the recruitment process is completely determined by $t\textsuperscript{intro}_i$, such that $\matr{\Psi}_{i, t}$ is:

\[
\begin{blockarray}{ccccc}
  \text{Not recruited} & \text{Uninfected} & \text{Infected} & \text{Dead} & \\
    u_{i, t+1} = 1 & u_{i, t+1} = 2 & u_{i, t+1} = 3 & u_{i, t+1} = 4 \\
\begin{block}{(cccc)c}
  1 - I_{(t = (t\textsuperscript{intro}_i - 1))} & 0 & I_{(t = (t\textsuperscript{intro}_i - 1))} & 0 & \begin{tabular}{@{}c@{}}Not recruited \\ $u_{i, t} = 1$ \end{tabular} \\
  0 & \phi^-_{i, t} (1 - \eta^+_t) & \phi^-_{i, t} \eta^+_t & 1 - \phi^-_{i, t} & \begin{tabular}{@{}c@{}}Uninfected \\ $u_{i, t} = 2$ \end{tabular} \\
  0 & \phi^+_{i, t} \eta^-_t & \phi^+_{i, t} (1 - \eta^-_t) & 1 - \phi^+_{i, t} & \begin{tabular}{@{}c@{}}Infected \\ $u_{i, t} = 3$ \end{tabular} \\
  0 & 0 & 0 & 1 & \begin{tabular}{@{}c@{}}Dead \\ $u_{i, t} = 4$ \end{tabular} \\
\end{block}
\end{blockarray}
 \]

\noindent where  $I_{(t = (t\textsuperscript{intro}_i - 1))}$ is an indicator function equal to one when $t$ is equal to $t\textsuperscript{intro}_i - 1$.

An imaginary first primary period augments the collection of primary periods when introductions took place and when surveys occurred, and was set to occur one week before the initial introductions into each population. 
For this augmented period $t = 1$, we assume that all individuals are in the ``not-recruited'' class, $u_{i, 1} = 1$ for $i = 1, ..., M$ (Figure \ref{hmm-structure}). 
Given that neither lake was known to contain adult frogs immediately prior to introduction, this is potentially a fair assumption, but in case the assumption was violated and adults were present prior to the introduction, we include a time-varying adjustment into the recruitment model (described below).

The number of primary periods was not uniform across survey years, and as such, the time between primary periods was also heterogeneous, complicating the interpretation of transition probabilities. 
Thus, at both sites, we used augmented primary periods to increase the regularity of time intervals between primary periods. 
These were chosen to occur when primary periods might have occurred, e.g., not during the winter months when the lakes are snow-covered, using a statistical model of when surveys are conducted. 
At each site, we fit a binomial generalized additive model with thin plate spline smoothers for day-of-year and daily snow depth at Dana Meadows, where the response variable was zero or one for each day, with one indicating that a survey took place. 
These models were fit using the \code{gam} function in the mgcv package in the R programming language [@r-lang; @wood2017]. 
Then, we predicted new values from these models, with the goal of augmenting the set of primary periods for each year at each site until all years except the first and last had the same number of primary periods (the maximum number of primary periods within one year at each site: ten at Alpine and eight at Subalpine). 
Augmented primary periods were only accepted if there was no primary period within five days (before and after) of the proposed augmented primary period. 
The end result is a set of primary periods that are more uniformly distributed in time than the original collection of empirical primary periods, though these augmented primary periods are not associated with real surveys.

### Parameter model

#### Detection probabilities

Among uninfected adult frogs, we assume that the probability of detection varies with survey air temperature [@sinsch1984thermal], so that

$$\text{logit}(p_{i, j}^-) = \alpha^{(p)} + \beta^{(p, x)} x_{j},$$

\noindent where $\alpha^{(p)}$ is an intercept parameter, $\beta^{(p, x)}$ is the effect of survey air temperature on detection probability, and $x_j$ is the survey air temperature for survey $j$, for all $i, j$. 
Among infected adult frogs, the detection model was expanded to allow for an adjustment to account for being infected, and a further adjustment to deal with variation due to Bd load:

$$\text{logit}(p_{i, j}^+) = \alpha^{(p)} + \beta^{(p, x)} x_{j} + \beta^{(p, +)} + \beta^{(p, z)} z_{i, t_j},$$

\noindent where $\beta^{(p, +)}$ is the adjustment on the intercept for infected adults, and $\beta^{(p, z)}$ is a coefficient for the log Bd load $z_{i, t_j}$ of individual $i$ in primary period $t$ containing survey $j$, for all $i, j$.

#### Recruitment probabilities

The recruitment model was designed to account for annual variation in Bd loads, whether primary periods spanned years, and winter severity. 
For the probability of entering the population between primary period $t$ and $t+1$, we have:

$$\text{logit}(\lambda_t) = \alpha^{(\lambda)} + \beta^{(\lambda, w)} w_t + \beta^{(\lambda, s)} s_{y_t} + \beta^{(\lambda, 1)} I_{(t = 1)} + \epsilon^{(\lambda)}_{y_t},$$

\noindent where $\alpha^{(\lambda)}$ is an intercept term, and the effect of an overwinter transition is represented as $\beta^{(\lambda, w)}$, with $w_t$ as a binary indicator of whether a transition from period $t$ to $t + 1$ spans a winter (or equivalently, two years).
The effect of the previous winter's severity is $\beta^{(\lambda, s)}$, where $s_{y_t}$ is previous winter's severity. 
The parameter $\beta^{(\lambda, 1)}$ is an adjustment for the recruitment probability after the first imaginary primary period, which could account for undetected individuals present prior to introduction.
Finally, $\epsilon^{(\lambda)}_{y_t}$ is an adjustment to account for extra annual variation. 
The probability that an individual is infected, conditional on recruitment is modeled as a function of expected Bd load among infected adults:

$$\text{logit}(\gamma_t) = \alpha^{(\gamma)} + \beta^{(\gamma, \mu)} \mu_{y_t} + \epsilon^{(\gamma)}_{y_t},$$

\noindent where $\alpha^{(\gamma)}$ is an intercept, $\beta^{(\gamma, \mu)}$ represents the effect of mean Bd load among infected adults in the year containing primary period $t$ (denoted $\mu_{y_t}$), and $\epsilon^{(\gamma)}_{y_t}$ is a year-specific adjustment.

#### Survival probabilities

Survival of uninfected adults was modeled as a function of whether a transition spanned an overwinter period and winter severity:

$$\text{logit}(\phi_{i, t}^-) = \alpha^{(\phi^-)} + \beta^{(\phi^-, w)} w_t + \beta^{(\phi^-, s)} s_{y_t} + \epsilon^{(\phi^-)}_{y_t},$$

\noindent where $\alpha^{(\phi^-)}$ is an intercept parameter, $\beta^{(\phi^-, w)}$ is an adjustment for overwinter transitions, $\beta^{(\phi^-, s)}$ is a coefficient for winter severity, and $\epsilon^{(\phi^-)}_{y_t}$ is a year-specific adjustment. 
The survival probabilities for infected adults were similarly modeled, but with additional effects of individual Bd load:

$$\text{logit}(\phi_{i, t}^+) = \alpha^{(\phi^+)} + \beta^{(\phi^+, w)} w_t + \beta^{(\phi^+, s)} s_{y_t} + \beta^{(\phi^+, z)} z_{i, t} + \epsilon^{(\phi^+)}_{y_t},$$

\noindent where $z_{i, t}$ is the log transformed Bd load of individual $i$ during primary period $t$, $\beta^{(\phi^+, z)}$ is a coefficient for Bd load, and the remainder of parameters are defined using the same notation as for the survival of uninfected adults.

#### Loss and gain of infection probabilities

The probability that an infected adult loses infection was modeled as a function of mean Bd load in the infected population, and whether a transition occurred from one year to the next:

$$\text{logit}(\eta_t^-) = \alpha^{(\eta^-)} + \beta^{(\eta^-, \mu)} \mu_{y_t} + \epsilon^{(\eta^-)}_{y_t},$$

\noindent where $\alpha^{(\eta^-)}$ is an intercept, $\beta^{(\eta^-, \mu)}$ is the effect of expected Bd load among infected adults, and  $\epsilon^{(\eta^-)}_{y_t}$ is a year-specific adjustment. 
Transitions from the uninfected to infected class were modeled similarly:

$$\text{logit}(\eta_t^+) = \alpha^{(\eta^+)} + \beta^{(\eta^+, \mu)} \mu_{y_t} + \epsilon^{(\eta^+)}_{y_t}$$

\noindent where parameter notation conventions and definitions match those for transitions from the infected to uninfected classes.

#### Bd loads

The fact that individuals are imperfectly detected, and that occasionally, individuals are detected but no infection data are collected presents a challenge for including individual-level Bd loads in the model. 
Within the infected population, Bd loads are partially observed when individuals are captured and swabs are collected. 
We used a normal distribution to represent (potential) log Bd loads:

$$[\matr{Z} | \mu, \sigma] = \prod_{i = 1}^M \prod_{t = 1}^{n_t} \text{Normal}(z_{i, t} | \mu_{y_t}, \sigma)$$

\noindent When the load of individual $i$ during primary period $t$ was observed, this specifies a likelihood. 
Otherwise, this specifies a prior distribution for the potential log load of individual $i$ on primary period $t$, conditional on infection. 
The expected value of Bd load among infected adults was assumed to vary among years, and potentially vary as a function of winter severity:

$$[\mu | \alpha^{(\mu)}, \beta^{(\mu)}, \sigma^{(\mu)}] = \prod_{y = 1}^{n_y} \text{Normal}(\mu_{y} | \alpha^{(\mu)} + \beta^{(\mu)} s_y, \sigma^{(\mu)})$$

#### Prior distributions

Based on knowledge of the observation process, we expected that most captures resulted in a non-missing swab result, leading to the specification for the prior on $\delta$ as $[\delta] = \text{Beta}(\delta | 9, 1)$.
Annual adjustments were modeled using zero-mean normal distributions with unknown standard deviations specific to the process of interest, e.g., for the probability of entering the population: $[\epsilon^{(\lambda)}] = \prod_{y = 1}^{n_y} \text{Normal}(\epsilon^{(\lambda)}_y | 0, \sigma^{(\epsilon^{(\lambda)})})$. 
Standard deviation parameters were given half normal priors i.e., $[\sigma^{(\epsilon^{(\lambda)})}] = \text{Normal}_+(\sigma^{(\epsilon^{(\lambda)})} | 0, 1.5)$, and all remaining parameters were given $\text{Normal}(0, 1.5)$ priors.
The full factorization of the joint distribution of data and parameters is provided in \hyperref[joint-distribution]{Appendix S1: Joint distribution factorization}.

#### Parameter estimation

We implemented the model in Stan, a probabilistic programming language, and sampled from the approximate posterior distribution of parameters with automatic differentiation variational inference [@carpenter2016stan; @kucukelbir2015automatic].
We used the forward algorithm [@zucchini2016hidden] to circumvent the need to sample from the discrete space of true individual states (see \hyperref[forward-algorithm]{Appendix S2: Forward algorithm description} for details).
Variational inference uses a simple family of distributions as an approximation of the posterior distribution, optimizing parameters of this simpler approximation (the variational distribution) to minimize the Kullback-Leibler divergence between the true posterior and the variational distribution. 
A variational approximation was necessary for the models to run in a reasonable amount of time ($\approx 2-4$ hours) on finite resources (r4.2xlarge EC2 instances on Amazon Web Services, with 8 virtual CPUs and 61 GiB RAM). 
To verify our model implementation, we simulated data with known parameters from the model to evaluate parameter recovery, and check that parameters were identifiable. 
These simulations allowed us to identify a lack of identifiability in an earlier model specification which included multiplicative interaction terms between winter severity and Bd load for both the probability of entering the population, and the probability that new adults were infected conditional on recruitment.

All code and data necessary to reproduce the analysis and manuscript is publicly available at [https://www.github.com/snarl1/sierra-reintroduction-cmr](https://www.github.com/snarl1/sierra-reintroduction-cmr) [@joseph_2018]. 
The workflow is wrapped into GNU Make command [@Stallman2004], the manuscript is written in R Markdown [@rmarkdown], and we used the R programming language [@r-lang] with the assertthat, ggrepel, ggridges, ggthemes, lubridate, mgcv, patchwork, reshape2, rstan, and tidyverse packages to facilitate data processing, model fitting, and visualization [@assertthat; @ggrepel; @ggridges; @ggthemes; @lubridate; @mgcv; @patchwork; @reshape2; @rstan; @tidyverse].
The computational environment with these dependencies is containerized via Docker, and the Dockerfile for the image exists in the GitHub repository for this project [@boettiger2015introduction].

# Results

## Detection probabilities

```{r load-detection-results}
detection_probs <- read_csv('out/detection-probs.csv')
detection_effs <- read_csv('out/detection-effs.csv')
```

>>>>

---
title: 'Appendix S2: Forward algorithm description'
subtitle: 'From: **Disease and climate effects on individuals drive post-reintroduction population dynamics of an endangered amphibian** in the journal Ecosphere'
author: 'Maxwell B. Joseph & Roland A. Knapp'
output: 
  pdf_document
header-includes:
  - \newcommand{\matr}[1]{\mathbf{#1}}
---

Parameter estimation for this model is made somewhat difficult by the presence of discrete parameters (hidden states).
We address this issue by using the forward algorithm, which does not require sampling from the discrete state space, to compute the joint probability of hidden states and observations [@zucchini2016hidden].
To describe this algorithm, we first consider the case of one individual. 
We would like to compute $[u_{i, 1:n_t}, o_{i, 1:n_j} | ...]$ (suppressing dependence on detection and transition parameters for compactness) for the individual with state and capture history shown in Figure 1.  
We can factor this joint probability as follows:
